
## ADVENTIST UNIVERSITY OF CENTRAL AFRICA_RWANDA
<center><H3> R for DATA SCIENCE</H3> </center>
# STUDENT NAMES_ID: Erneste UWIMPUHWE_101414
# LECTURER:
# Dr.Pacifique Nizeyimana
# DATE 17/10/2025
<h3>class one Task</h3>

# DAY1_PRACTICAL_ACTIVITIES_IN_R
# steps of installing R 

* Download R from the CRAN website based on your operating system
* Run the installer and follow default setup instructions 
* Open the R console to verify successful installation  
* Optionally, install RStudio for a better interface 
* Test the setup by running a simple R command like 2+2 

# Simple calculator in R
```{r}
2+7    #PERORMING ADDITION
200*14 #PERFORMING MULTIPLICATION
34/23 # PERFORMING DIVISION
18-89 #PERFORMING SUBSTARCTION

```

# 3 Data structure (Vectors, Factors, Lists, Data frame, matrix and arrays)
# 3.1 Vectors

```{r}
x <- c(0.5, 0.6)
age<-c(20,35,32,29)
summary(age)
```
# CALCULATING MEAN OF VECTOR
```{r}
age<-c(35,24,18,24)
mean(age)
```
# 3.2 Stucture of LOGICAL
```{r}
x <- c(TRUE, FALSE)
x <- c(T, F)
```
# 3.3 Stucture of Character
```{r}
x <- c("a", "b", "c" )
class<-c("M","F","F","M")
```

# 3.4 Stucture of Integer
```{r}

x <- 9:29
```
# 3.6 Stucture of Complex
```{r}
x <- c(1+0i, 2+4i)
```

# Matrix and operation within matrices
```{r}
x<- matrix(1:6, nrow = 2, ncol = 3)
x
```
```{r}
(y<-matrix(1:6, nrow = 3, ncol = 2))
```
```{r}
x%*%y
```
# 3.7 list
```{r}
x <- list(1, "a", TRUE, 1 + 4i)
x
```
#Factor
```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
```
# Create a vector with NAs in it
```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
```
```{r}
x <- c(1, 2, 4, "NA", 5)
bad <- is.na(x)
print(bad)
```
```{r}
x[!bad]
```
# if there are multiple R objects and you want to take the subset with no missingvalues in any of those object
```{r}
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x, y)
good
```
# Coercion
```{r}
sum(c(TRUE,TRUE,FALSE,FALSE,FALSE))#If a vector only contains logical and numbers, R will convert the logical to numbers, Every true becomes a 1, and every FALSE becomes 0
```
# Creating data frame
```{r}
studentID<-c(1,2,3,4,5)
math_score<-c(12,17,10,9,NA)
gender<-c("M","F","M","M","F")
it_score<-c(13,18,11,10,19)
scoredata<-data.frame(studentID,gender,math_score,it_score)
View(scoredata)
```
#Create data from keyboard
```{r}
data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric(0))
data_class2<-edit(data_class2)
```

```{r}
data_class <- read.table("E:/My documents-erneste/AUCA/R programming/class_data.txt")
variable.names(data_class)
```
```{r}
head(data_class)
```

```{r}
tail(data_class)
```
```{r}
data_class[10:20,]
```
```{r}
summary(data_class$WEIGHT)
```

```{r}
length(data_class$WEIGHT)
```
```{r}
data_class[,-1]
```

```{r}
data() # list of datasets currently available
data("airquality")
variable.names(airquality)
```
```{r}
str(airquality)
```


# Relationship between a categorical and a continuous variable
```{r}
data("iris")
head(iris)
```
#Exploration data analysis
##Examine numerical variable using common summary
```{r}
data_class
```
```{r}
summary(data_class$)
```
```{r}
str(data_class)
```
#Summary Statistics
```{r}
# Convert a 'character' column to 'numeric'
data_class$V1 <- as.integer(data_class$V1)
data_class$V2 <- as.integer(data_class$V2)

```
```{r}
str(data_class)
```

```{r}
summary(data_class$V2)
```
```{r}
apply(data_class,2,mean)
```
```{r}
apply(data_class,2,sd)
```
```{r}
c(mean(data_class$V1),sd(data_class$V2))
```
```{r}
c(mean(data_class$WEIGHT),sd(data_class$WEIGHT))
```


```{r}
mean(iris$Petal.Length)
mean(iris$Petal.Length[iris$Species=="setosa"])
mean(iris$Petal.Length[iris$Species=="versicolor"])
mean(iris$Petal.Length[iris$Species=="virginica"])
by(iris$Petal.Length,iris$Species,mean)
by(iris$Petal.Length,iris$Species,summary)
```
# Visualize the differences in continuous variables between categories using Box-and- whisker plot
```{r}
boxplot(iris$Sepal.Length~iris$Species,data=iris, main="Comparison")
```
# Relationship between two categorical variables
```{r}
ucba<-data.frame(UCBAdmissions)
head(ucba)
```
```{r}
cross<-xtabs(Freq~Gender+Admit,data=ucba)
(cross<-xtabs(Freq~Gender+Admit,data=ucba))
```
```{r}
prop.table(cross,2)
```

#Simpson’s paradox
```{r}
cross2<-xtabs(Freq~Gender+Admit,data=ucba[ucba$Dept=="A",])
prop.table(cross2,1)
```
# Bar chat
```{r}
dat <- read.table(text ="ProdA ProdB ProdC ProdD
1 110 50 60 70
2 120 50 80 65", header= TRUE)
barplot(as.matrix(dat),beside=FALSE,col=c("Red","green"))
```
# barplot(as.matrix(dat),beside=TRUE,col=c("gold3","red"))
```{r}
dat <- read.table(text = "A B C D E F G
1 10 80 30 90 70 60 90
2 20 50 70 50 40 10 40
3 60 80 80 60 60 30 160
4 20 40 70 80 20 10 70", header = TRUE)
barplot(as.matrix(dat))
```
# SCATTER PLOT

```{r}
plot(V2~V1,data=data_class)
```
```{r}
plot(data_class$V1,data_class$V2)
```

#Linear Models
```{r}

plot(V2~V1,data=data_class)
abline(lm(V2~V1,data=data_class)$coefficient)
```
```{r}
data_lm<-lm(V2~V1,data=data_class)
summary(data_lm)
```

```{r}
hist(data_class$V1)
```
```{r}
hist(data_class$V2,col = "blue")
```
```{r}
plot(V2~V1,data=data_class)
abline(lm(V2~V1,data=data_class)$coefficient)
lines(lowess(data_class$V1,data_class$WV2),col="blue")
```
```{r}
cor(data_class$V1,data_class$V2)
```
```{r}
attach(data_class)
(BMI<-V2/(V1/100)^2)
```

```{r}
(BMI<-round(V2/(V1/100)^2,digit=2))
```
```{r}
head(cbind(data_class,BMI),n=5)
```
```{r}
new_data_class<-cbind(data_class,BMI)
tail(cbind(data_class,BMI),n=10)
```

```{r}
new_data_class<-cbind(data_class,BMI)
tail(cbind(data_class,BMI),n=10)
```

```{r}
summary(BMI)
```
```{r}
stem(BMI,scale=2)
```

